---
title: "ShootingRmd"
author: "A. K. Attri"
date: "2022-12-27"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```
## Importing NYPD Shooting Dataset

I will start by reading in the data from the csv files.

```{r set_nypd_shooting_data_url}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
```

Let's read in the data and explore it
```{r import_data, message=FALSE}
nypd_shooting_data <- read_csv(url)
nypd_shooting_data
```

## Tydying Up and Transforming NYPD Shooting Dataset
Now that I have explored the data, I would tidy it and store total shooting cases per year and total murder cases per year in a separate data frame. I will only keep year and count data and remove all other columns from it.

```{r tidy_cases_data}
## Get total shooting cases per year
total_shooting_cases_per_year <- nypd_shooting_data %>% count(OCCUR_DATE, name = "value") %>%
mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
mutate(month = format(OCCUR_DATE, "%m"), year = format(OCCUR_DATE, "%Y")) %>%
group_by(year) %>%
summarise(total_shooting_cases = sum(value))
total_shooting_cases_per_year

## On analyzing the total shooting cases per year we found that max recorded cases were 2055 in the year 2006 and min cases were 958 in the year 2018. On avg 1600 cases were recorded per year from 2006 to 2021
summary(total_shooting_cases_per_year)

## Get total murder cases per year
total_murder_cases_per_year <- nypd_shooting_data %>% filter(STATISTICAL_MURDER_FLAG == TRUE) %>% count(OCCUR_DATE, name = "value") %>%
mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
mutate(month = format(OCCUR_DATE, "%m"), year = format(OCCUR_DATE, "%Y")) %>%
group_by(year) %>%
summarise(total_murder_cases = sum(value))
total_murder_cases_per_year

## On analyzing the total murder cases per year we found that max recorded cases were 445 in the year 2006 and min cases were 174 in the year 2017. On avg 308 cases were recorded per year from 2006 to 2021
summary(total_murder_cases_per_year)
```

I will combine both these tables on year so that I can plot it and visualize the correlation between total shooting cases and murder cases. Also, I will see how cases increased/decreased over the years.

```{r transform_data}
combined_cases_data <- inner_join(total_shooting_cases_per_year, total_murder_cases_per_year, by="year")
combined_cases_data$year <- as.numeric(combined_cases_data$year)

combined_cases_data

```

## Including Plots

I have ploted the total shooting cases and murder cases on the graph. 

```{r plot, echo=FALSE}
combined_cases_data %>%
ggplot(aes(x=year, y=total_shooting_cases)) +
geom_line(aes(color="total_shooting_cases")) +
geom_point(aes(color = "total_shooting_cases")) +
geom_line(aes(y = total_murder_cases, color="total_murder_cases")) +
geom_point(aes(y = total_murder_cases, color = "total_murder_cases")) +
theme(legend.position = "bottom",
axis.text.x = element_text(angle = 90)) +
scale_x_continuous(breaks = scales::pretty_breaks(n = 20)) +
scale_y_continuous(breaks = scales::pretty_breaks(n = 20)) +
labs(title = "NYPD Shooting Cases", y = NULL)
```

## Analysis
From the graph we can see that cases were declining from year 2006 till 2019 but then again shot up to the levels of 2006 in 2020 and 2021. 
Interesting to note that this was the time of covid epidemic as well during which shooting cases increased.
Also, there seems to be a linear relationship between shooting cases and murders which is expected.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Modeling Data
```{r}
mod <- lm(total_murder_cases ~ total_shooting_cases, data = combined_cases_data)
summary(mod)

combined_cases_data_pred <- combined_cases_data %>% mutate(pred = predict(mod))
combined_cases_data_pred

combined_cases_data_pred %>% ggplot() + 
     geom_point(aes(x = total_shooting_cases, y=total_murder_cases), color = "blue") +
     geom_point(aes(x=total_shooting_cases, y=pred), color = "red")
```

From the model we can see that murder cases follow shootings linearly and our linear model is able to predict murders from shootings with fair accuracy.

## Bias
I had a bias in mind that during covid shooting incidents will decrease due to lockdowns and restricted movements but that was proven wrong. Thin police resources due to focus on covid response and illness could be a source of increase in shooting incidents during Covid.    